Enabling Ambient Backscatter \\Using Low-Cost Software Defined Radio

### 2017-04-07 paper.tex
Added heatmap.
Worked on the RTL2832U section.

In this section we examine some background aspects that are essential to our
work. \cp{Put a simple intro to each section so that they don't look empty
before the first subsection.}

@@

\subsection{RTL2832U}
\label{sub:rtl2832}
\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
				\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q demodulator \textit{Raeltek RTL2832U} (left IC) and the tuner with integrated LNA \textit{Rafael Micro R820T/2} (right IC).}
\label{fig:receiver_arch}
\end{figure}
The Raeltek RTL2832U is a terrestrial digital video broadcast (DVB-T) demodulator IC, which is at the core of a wide range of USB-based DVB-T receiver dongles. One prominent feature of this chip is that it allows to retrieve the raw I/Q samples via USB. This is originally intended for the chip to work as a simple DAB/FM software receiver. Ham radio enthusiasts have combined their efforts to create a software driver (the \textit{librtlsdr}, cf. \cite{steve-m_librtlsdr}) that allows DVB-T dongles based on the RTL2832 to be converted into a low-cost wideband SDR receiver. The cost of traditional SDR's has been in the range of a few hundred to thousands of USD. Albeit much less powerful than the typical SDR, an RTL2832U-based DVB-T receiver can be bought with antenna for less than 10 USD.
An RTL-SDR does not only consist of the RTL2832U. The second essential component is the tuner, which allows the user to select the received frequency and outputs that signal at an intermediate frequency (IF). The IF can then be fed into the demodulator for digitalization. In the case of the used hardware this tuner is a \textit{Rafael Micro R820T/2} which offers a tuning range from 42 to 1002 MHz~\cite{rafael_r820t_2011}. It subsumes a low-noise amplifier (LNA) and filters. It presents a low noise figure of 3.5 dB and features 65 dBc of image rejection. Figure \ref{fig:receiver_arch} \cp{We need a much better photo here.}, shows a photograph of the RTL-SDR used in our work.

@@

The received signal can be interpreted as
\begin{equation}
	s_{IF}(t)=I(t) \cdot cos(\omega_{0}t) + Q(t) \cdot sin(\omega_{0}t)
\end{equation}     
This is multiplied with a cosine (and a sine as well) from a local oscillator.
\begin{multline}
        s_{IF}(t) \cdot s_L(t) = I(t) \cdot cos(\omega_{0}t) \cdot cos(\omega_{0}t)\\+ Q(t) \cdot sin(\omega_{0}t) \cdot cos(\omega_{0}t)
\end{multline}
With \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)} and \ensuremath{2\,sin(a)cos(b)=sin(a+b)+sin(a-b)} follows
\begin{multline}
        2 \cdot s_{IF}(t) \cdot s_L(t) = I(t) \cdot \bigl[1+cos(2\omega_{0}t)\bigr]\\+ Q(t) \cdot \bigl[sin(2\omega_0t)+sin(0)\bigr].
\end{multline}
One can see, that the interesting in-phase part (I) in this case is represented by a DC value after mixing. With a low-pass filter this DC value can be separated from the undesired rest. An analogous procedure is done with the quadrature part (Q) when mixing with a sine. \cp{We need to explain this more clearly and possibly more concisely.}

@@

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.8\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the self-made backscattering ground-plane antenna.}
\label{fig:s11}
\end{figure}
A backscattering antenna is the most important part of backscatter transmitter. Our antenna is optimized for usage with the television signal, which has its center frequency at 626 MHz. It is a ground plane antenna, which has been built up on a large unetched PCB plate. This rectangular PCB, with dimensions of 200 mm x 300 mm, is covered with a copper layer on one side and left blank on the other. A hole in the middle of the PCB serves for connecting a wire to a 50 \ensuremath{\Omega} micro strip transmission line. The wire, which acts as the vertical antenna element and has a length of 350 mm (roughly 0.73 \ensuremath{\lambda}), is soldered to the transmission line. The transmission line has been calculated for 626 MHz using an FR4 substrate with a thickness of 1.6 mm. According to micro strip line theory the length of the line changes the phase and the width of the line changes the impedance (cf. Chapter 3 in \cite{pozar_microwave_2011}). The design equations given in \cite{pozar_microwave_2011} are implemented in various programs e.g. KiCAD, which has been used for solving the design equations. The given parameters lead to a micro strip line width of 3 mm. An open-stub, attached directly at the juncture where the vertical antenna element is attached to the transmission line, is used for fine tuning of the input impedance. The stub length has been adjusted manually (shortened) while measuring the reflection coefficient with a vector network analyzer (VNA). Open-stub matching is discussed in Chapter 5 in \cite{pozar_microwave_2011}. Furthermore the resonance frequency has been tuned by cutting the vertical antenna element piece by piece while measuring with the VNA until \ensuremath{S_{\text{11}}} had the minimum close to the desired frequency (Figure~\ref{fig:s11}\cp{Can we replace all these figures by PDFs?}). In the plot one can also see, that the antenna is quite narrowband\cp{We should report something like the 3dB bandwidth rather than saying something non-specific like this}. For backscattering this is not necessarily a disadvantage, since everything that is within the operating frequency of the antenna is backscattered. With our narrowband antenna the undesired backscattering of out-of-band signals is reduced. With our approach we were able to achieve an input impedance at 634 MHz of \ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz \ensuremath{S_{\text{11}}} is still approximately -17 dB.

@@

During the scanning process	the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be justified by the fact, that DVB-T is specified up to 10 MHz bandwidth. And the local TV signal provider is to its own statements using DVB-T 2. The signal \ensuremath{|H|} is calculated as follows.
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\bigl( \Re\{ s_{\text{band,n}}(t) \} \bigr) \Biggr|
\end{equation}     
Where \ensuremath{N} is the total number of FFT bins accumulated over
the entire frequency sweep\cp{Please confirm I got this right}. The FFT
is only carried out over the real values, which are received from the
\textit{RTL-SDR}. \ensuremath{s_{\text{band,n}}(t)} is the signal in the
time domain of one band. There are as many samples taken in one band as
needed for an FFT of size 2048. No window is used when stitching
individual scans together. One can see, the effects of this a little bit, when taking a closer look at the plotted data. Rising and falling edges at the beginning of each band are visible. To get \ensuremath{|H_{tv,avg}|} one has to normalize everything and calculate the power. Before doing that a simple smoothing algorithm (FIR) is applied on \ensuremath{|H|}.
\begin{equation}
	|H_{tv,avg}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain
point in space. 
\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{./fig/haversine}
\caption{Signal strength fading against distance relative to the spot where the maximum signal strength has been captured (TV towers). 474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and 594 MHz belongs to a tower in Uppsala, Rickomberga. Distances have been calculated with the Haversine formula.}
\label{fig:haversine}
\end{figure}

>>>>

In this background section we introduce the necessary knowledge about sofware defined radios and the used hardware, the \textit{RTL-SDR}.

@@

\subsection{RTL2832U}
\label{sub:rtl2832}
 \begin{figure}[h]
 \centering
\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q demodulator
\textit{Raeltek RTL2832U} (left IC) and the tuner with integrated LNA
\textit{Rafael Micro R820T/2} (right IC).}
 \label{fig:receiver_arch}
\end{figure}
The \textit{Realtek RTL2832U} is a terrestrial digital video broadcast
(DVB-T) demodulator IC, which is the main part of a wide range of
USB-based DVB-T receiver dongles. One prominent feature of this chip is
that it allows to retrieve the raw I/Q samples via USB. This is
originally intended for the chip to work as a simple DAB/FM software
receiver. Ham radio enthusiasts have combined their efforts to create a
software driver (the \textit{librtlsdr}, cf. \cite{steve-m_librtlsdr})
that allows DVB-T dongles based on the \textit{RTL2832} to be converted
into low-cost wideband SDR receivers. The cost of traditional SDRs has
been in the range of a few hundred to thousands of USD. Albeit much less
powerful than a typical SDR, a \textit{RTL2832U}-based DVB-T receiver
can be bought with antenna for less than 10 USD.  A RTL-SDR does not
only consist of the \textit{RTL2832U}, but also a tuner. The tuner is an
essential component, as it allows the user to select the received
frequency and outputs that signal at an intermediate frequency (IF). The
IF can then be fed into the demodulator for digitalization. In the case
of the used hardware this tuner is a \textit{Rafael Micro R820T/2} which
offers a tuning range from 42 to 1002 MHz~\cite{rafael_r820t_2011}. It
subsumes a low-noise amplifier (LNA) and filters. It presents a low
noise figure of 3.5 dB and features 65 dBc of image rejection. Figure
\ref{fig:receiver_arch}, shows a photograph of the RTL-SDR used in our
work.

@@

The received signal can be interpreted as
\begin{equation}
	s_{IF}(t)=I(t) \cdot cos(\omega_{0}t) + Q(t) \cdot sin(\omega_{0}t)
\end{equation}     
This is multiplied with a cosine (and a sine as well) from a local oscillator.
\begin{multline}
        s_{IF}(t) \cdot s_L(t) = I(t) \cdot cos(\omega_{0}t) \cdot cos(\omega_{0}t)\\+ Q(t) \cdot sin(\omega_{0}t) \cdot cos(\omega_{0}t)
\end{multline}
With \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)} and \ensuremath{2\,sin(a)cos(b)=sin(a+b)+sin(a-b)} follows
\begin{multline}
        2 \cdot s_{IF}(t) \cdot s_L(t) = I(t) \cdot \bigl[1+cos(2\omega_{0}t)\bigr]\\+ Q(t) \cdot \bigl[sin(2\omega_0t)+sin(0)\bigr].
\end{multline}
One can see, that the interesting in-phase part (I) in this case is
represented by a DC value after mixing. With a low-pass filter this DC
value can be separated from the undesired rest. An analogous procedure
is done with the quadrature part (Q) when mixing with a sine.

@@

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the self-made backscattering ground-plane antenna.}
\label{fig:s11}
\end{figure}
A backscattering antenna is the most important part of backscatter transmitter. Our antenna is optimized for usage with the television signal, which has its center frequency at 626 MHz. It is a ground plane antenna, which has been built up on a large unetched PCB plate. This rectangular PCB, with dimensions of 200 mm x 300 mm, is covered with a copper layer on one side and left blank on the other. A hole in the middle of the PCB serves for connecting a wire to a 50 \ensuremath{\Omega} micro strip transmission line. The wire, which acts as the vertical antenna element and has a length of 350 mm (roughly 0.73 \ensuremath{\lambda}), is soldered to the transmission line. The transmission line has been calculated for 626 MHz using an FR4 substrate with a thickness of 1.6 mm. According to micro strip line theory the length of the line changes the phase and the width of the line changes the impedance (cf. Chapter 3 in \cite{pozar_microwave_2011}). The design equations given in \cite{pozar_microwave_2011} are implemented in various programs e.g. KiCAD, which has been used for solving the design equations. The given parameters lead to a micro strip line width of 3 mm. An open-stub, attached directly at the juncture where the vertical antenna element is attached to the transmission line, is used for fine tuning of the input impedance. The stub length has been adjusted manually (shortened) while measuring the reflection coefficient with a vector network analyzer (VNA). Open-stub matching is discussed in Chapter 5 in \cite{pozar_microwave_2011}. Furthermore the resonance frequency has been tuned by cutting the vertical antenna element piece by piece while measuring with the VNA until \ensuremath{S_{\text{11}}} had the minimum close to the desired frequency (Figure~\ref{fig:s11}\cp{Can we replace all these figures by PDFs?}). In the plot one can also see, that the antenna is quite narrowband\cp{We should report something like the 3dB bandwidth rather than saying something non-specific like this}. For backscattering this is not necessarily a disadvantage, since everything that is within the operating frequency of the antenna is backscattered. With our narrowband antenna the undesired backscattering of out-of-band signals is reduced. With our approach we were able to achieve an input impedance at 634 MHz of \ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz \ensuremath{S_{\text{11}}} is still approximately -17 dB.

@@

During the scanning process	the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be justified by the fact, that DVB-T is specified up to 10 MHz bandwidth. And the local TV signal provider is to its own statements using DVB-T 2. The signal \ensuremath{|H|} is calculated as follows.
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\bigl( \Re\{ s_{\text{band,n}}(t) \} \bigr) \Biggr|
\end{equation}     
Where \ensuremath{N} is the total number of FFT bins accumulated over
the entire frequency sweep\cp{Please confirm I got this right}. The FFT
is only carried out over the real values, which are received from the
\textit{RTL-SDR}. \ensuremath{s_{\text{band,n}}(t)} is the signal in the
time domain of one band. There are as many samples taken in one band as
needed for an FFT of size 2048. No window is used when stitching
individual scans together. One can see, the effects of this a little bit, when taking a closer look at the plotted data. Rising and falling edges at the beginning of each band are visible. To get \ensuremath{|H_{tv,avg}|} one has to normalize everything and calculate the power. Before doing that a simple smoothing algorithm (FIR) is applied on \ensuremath{|H|}.
\begin{equation}
	|H_{tv,avg}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain
point in space. 
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\caption{Signal strength fading against distance relative to the spot where the maximum signal strength has been captured (TV towers). 474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and 594 MHz belongs to a tower in Uppsala, Rickomberga. Distances have been calculated with the Haversine formula.}
\label{fig:haversine}
\end{figure}
